---
title: "Untitled"
author: "J Lacasa"
date: "2026-06-24"
output: html_document
---

```{r}
library(tidyverse)
library(emmeans)
library(lme4)
```


### Applied case II -- RCBD

- Field experiment studying the effect of potassium on corn yield.
- One treatment factor: Potassium fertilizer (one-way treatment structure).
- Randomized Complete Block Design with 4 repetitions (design structure).
















```{r}
url <- "https://raw.githubusercontent.com/stat720/summer2026/refs/heads/main/data/cochrancox_kfert.csv"

df2 <- read.csv(url) |> 
  transmute(K2O_lbac = as.factor(K2O_lbac), rep = as.factor(rep), yield)
```

Because this is an RCBD, the model is 

$$y_{ij} = \mu_0 + T_i + b_j + \varepsilon_{ij}, \\  \varepsilon_{ij} \sim N(0, \sigma^2).$$

**Blocks: fixed of random?** 

- The assumptions behind $b_j$ vary: 
- Fundamentals of mixed models 
  - Assumption behind blocks as fixed effects: there is a 'true' block effect out there. 
  - Assumption behind blocks as random effects: 
- Confidence intervals of the means differ depending on the model:  
  - Blocks as fixed: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$
  - Blocks as random: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$ 
- Confidence intervals of the means differences don't differ depending on the model:  
  - Blocks as fixed/as random: $\hat\mu \pm t \cdot \sqrt{\frac{2 \sigma_{\varepsilon}^2 }{b}}$  


Blocks as fixed: 

- very few blocks 
- blocks were very different 

```{r}
m_fixed <- lm(yield ~ K2O_lbac + rep, data = df2)

summary(m_fixed)

emmeans(m_fixed, ~ rep)

emmeans(m_fixed, ~ K2O_lbac) |> 
  as.data.frame() |> 
  ggplot(aes(K2O_lbac , emmean))+
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0)+
  geom_point()
emmeans(m_fixed, ~ K2O_lbac, contr = list(c(1, -1, 0, 0, 0), 
                                          c(1, 0, -1, 0, 0)))
```

Blocks as random: 

$$b_j \sim N(0, \sigma^2_b)$$


```{r}
m_random <- lmer(yield ~ K2O_lbac + (1|rep), data = df2)

emmeans(m_random, ~ K2O_lbac) |> 
  as.data.frame() |> 
  ggplot(aes(K2O_lbac , emmean))+
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0)+
  geom_point()


emmeans(m_random, ~ K2O_lbac, contr = list(c(1, -1, 0, 0, 0)))
```

Things to highlight: 

- Under an RCBD, se(mean) $\neq$ se(mean difference). 
- Standard errors of the means differ depending on the model:  
  - Blocks as fixed: $\sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$
  - Blocks as random: $\sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$ 
- Standard errors of the means differences don't differ depending on the model:  
  - Blocks as fixed/as random: $\sqrt{\frac{2 \sigma_{\varepsilon}^2 }{b}}$  


B = Fixed 
```{r}
sigma2_fixed <- sigma(m_fixed)^2
```

B = Random
```{r}
sigma2_random <- sigma(m_random)^2
sigma2b_random <- VarCorr(m_random)$rep[[1]]
```

SE MEAN (B = FIXED)
$$\sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$$

```{r}
#means
sqrt(sigma2_fixed/3)

#diff
sqrt(2*sigma2_fixed/3)
```
SE MEAN (B= RANDOM)
$$\sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$$ 
```{r}
#means
sqrt((sigma2_random + sigma2b_random)/3)

#diff
sqrt(2*(sigma2_random)/3)
```

## Applied case III -- split-plot design 

- Designed experiment of barley (70 levels) with fungicide (2 levels) treatments. 
- Two treatment factors: barley (70 levels) and fungicide (2 levels). Two-way factorial treatment structure.
- Split-plot design in an RCBD: Fungicide in the whole plot and genotype in the subplot.  














```{r}
agridat::durban.splitplot
df3 <- durban.splitplot

df3 |> #filter(fung == "F2", block == "B1") |> 
  ggplot(aes(bed, row))+
  geom_tile(aes(fill = fung))+
  coord_fixed()
```


Because this is a split-plot design, the model is 

$$y_{ij} = \mu_0 + F_i + G_j + (F\times G)_{ij} + b_k + wp_{i(k)} + \varepsilon_{ij}, \\
wp_{i(k)}, \sim N(0, \sigma_{wp}^2) \\ \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon).$$

- Like before, we can choose either fixed or random for $b_k$. 
- But now, the se(mean) depend on the level we are looking.  

For differences at the whole-plot level (fung): $se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon + g \ \sigma^2_{wp})}{b \cdot g}}$. 
For differences at the split-plot level (gen): $se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon)}{b \cdot f}}$. 

```{r message=F}
m3 <- lmer(yield ~ gen*fung + (1|block/fung), data = df3)


m3 <- lmer(yield ~ gen+ fung +gen:fung + (1|block) + (1|block:fung), data = df3)


emmeans(m3, ~fung, contr = list(c(1, -1)))$contr
```


```{r}
emmeans(m3, ~gen, contr = list(c(1, -1, rep(0, 68))))$contr
```


For differences at the whole-plot level (fung): $$se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon + g \ \sigma^2_{wp})}{b \cdot g}}$$

For differences at the split-plot level (gen): $$se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon)}{b \cdot f}}$$ 
```{r}
vc <- VarCorr(m3)
sigma2_b <- vc$block[1]
sigma2_wp <- vc$`block:fung`[1]
sigma2_epsilon <- sigma(m3)^2
n_block <- n_distinct(df3$block)
n_fung <- n_distinct(df3$fung)
n_gen <- n_distinct(df3$gen)

#se fung difference 
sqrt( 2*(sigma2_epsilon + (n_gen*sigma2_wp)) / (n_block *n_gen) )

#se gen difference 
sqrt( 2*(sigma2_epsilon ) / (n_block *n_fung) )
```

```{r}
test.t.fung <-  0.548 / 8.63  

(1 - pt(test.t.fung, df = 3))*2
```
```{r}

(pt(-1.085, df = 414))*2

  ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
    stat_function(fun = dt, args = list(df = 414), color = "darkgreen", size = 1) +
    labs(title = "Student's t-Distribution (df = 414)",
         x = "t Value",
         y = "Density") +
  geom_point(aes(x= -1.085, y=0), shape =21, fill = "gold", size = 3)+
    theme_minimal()
```






