---
title: "Crossover designs -- in class"
author: "J Lacasa"
date: "2026-07-14"
output: html_document
---

```{r warning=FALSE, message=FALSE}
library(lme4)
library(tidyverse)
library(emmeans)
```



## Example 1 

**Studying the effect of 2 treatments**

This study was published in [Grizzle (1965)](https://www.jstor.org/stable/2528104?seq=1). 

This study studied the effect of 2 drugs (A and B) on humans. 

First, 14 persons were randomly assigned one of two drugs, and the trait of study was measured at the end of the treatment duration. 

Then, all persons were given the other drug they hadn't taken yet. 

The model can be described as 

$$y_{ijkl} = \mu + T_i + P_j + S_k + u_{lk} + \varepsilon_{ijkl}$$

```{r message=FALSE, warning=FALSE}
df <- read.csv("https://raw.githubusercontent.com/stat720/summer2026/refs/heads/main/data/grizzle_messydata.csv")
df$Period <- as.factor(df$Period)
df$Person <- as.factor(df$Person)

m <- lmer(Y ~ Seq + Trt + Period + (1|Seq:Person) , data = df)

car::Anova(m, type = 3, test.statistic = "F")
```

```{r}
emmeans(m, ~ Trt, contr = list(c(1, -1)))
```

```{r message=FALSE, warning=FALSE}
plot(emmeans(m, ~ Trt, contr = list(c(1, -1))))
```

```{r}
unique(df |> dplyr::select(Person, Seq))
```

## Example 2  

**Studying the effect of 3 treatments**

6 cows were used to test 3 different treatments in a crossover design.  

- Repeated measures. Now we are assuming compound symmetry (a specific type of correlation function). 
- How does carryover affect power? 


```{r message=FALSE, warning=FALSE}
url <- "https://raw.githubusercontent.com/stat720/summer2025/refs/heads/main/data/cow_prod_cox.csv"
df <- read.csv(url)
df
unique(df$Sequence)
```

The model can be described as 

$$y_{ijkl} = \mu + T_i + P_j + S_k + u_{lk} + \varepsilon_{ijkl}$$
```{r message=FALSE, warning=FALSE}
m_milk.nocarryover <- lmer(milk_production ~ Sequence + Period + Treatment + (1|Sequence:Cow),
               data = df)
emmeans(m_milk.nocarryover, ~ Treatment )
emmeans(m_milk.nocarryover, ~ Treatment , 
        contr = list(c(1, -1, 0), 
                     c(1, 0, -1), 
                     c(0, 1, -1)))
```

Another model can be described as 

$$y_{ijklm} = \mu + T_i + P_j + S_k + C_m + u_{lk} + \varepsilon_{ijklm}$$


```{r}
m_milk.carryover <- lmer(milk_production ~ Sequence + Period + Treatment + Prior_trt + (1|Sequence:Cow),
               data = df)

emmeans(m_milk.carryover, ~ Treatment )

emmeans(m_milk.carryover, ~ Treatment , 
        contr = list(c(1, -1, 0), 
                     c(1, 0, -1), 
                     c(0, 1, -1)))

```

