# Power analysis II
July 10th, 2025  

## Announcements

- HW 3 due tomorrow 

## What are simulations? 

A lot the concepts we're learning in STAT 720 are based on asymptotic properties, or what would happen, *generally speaking*. 
For example, a 95% confidence interval will include the true value only 95% of the times. That means that 5% of the times, it will not. 
Likewise, a hypothesis test with an $\alpha=0.05$ will incorrectly reject the null hypothesis 5% $(=\alpha)$ of the times. 
That hypothesis test will also have an associated $\beta$, that will depend on factors like sample size and experiment design.  
If we wish to study how a design would work, we should test it many many times and not just once. 

Simulation studies are helpful to evaluate how a method performs, generally speaking. 

### Simulation demo  

**1. Open the required packages** 
```{r message=FALSE, warning=FALSE}
library(tidyverse) # data wrangling & data viz  
library(lme4) # model fitting
library(emmeans) # marginal means 
library(latex2exp) # math notation for plots
```

**2. Set the "true" values** 
```{r}
# set the true state
b0.truth <- 2 # true intercept value
b1.truth <- 4 # true slope value 
sigma.truth <- 1 # true variance

# create predictor
x <- seq(1, 60, by = 7)
```

**Demo: what happens if we simulate data based on known true values for the parameters, and then try to estimate the parameters again?**  
Most likely, the confidence intervals will contain the true value. Let's try it once: 

```{r}
# generate "fake data" based on the mean and some random error 
set.seed(22)
random_error <- rnorm(length(x), 0, sigma.truth)
y <- b0.truth + x*b1.truth + random_error
  
# fit the model to the data
m <- lm(y ~ x)

# 95% confidence intervals
confint(m)

between(b0.truth, confint(m)[1,1], confint(m)[1,2])
between(b1.truth, confint(m)[2,1], confint(m)[2,2])

```

**3. Do that 1000 times**  
The test above worked - the 95% CIs included the true values. Based on our knowledge though, we would expect some error rate. Let's check out what happens if we repeat this 1000 times:  

```{r eval=FALSE, echo=FALSE}
# set the number of simulations and create the space to store the estimates
n_sims <- 1000
b0.hat <- numeric(n_sims)
b1.hat <- numeric(n_sims)
b0.hat.calib<- numeric(n_sims)
b1.hat.calib<- numeric(n_sims)

set.seed(42)
for (i in 1:n_sims){
  # create random errors (these change from iteration to iteration)
  random_error <- rnorm(length(x), 0, sigma.truth)
  y <- b0.truth + x*b1.truth + random_error
  
  # fit the model to the (synthetic) data
  m <- lm(y ~ x)
  
  # store estimate
  b0.hat[i] <- coef(m)[1]
  b1.hat[i] <- coef(m)[2]
  
  # check if the confidence intervals include the true value
  b0.hat.calib[i] <- between(b0.truth, confint(m)[1,1], confint(m)[1,2])
  b1.hat.calib[i] <- between(b1.truth, confint(m)[2,1], confint(m)[2,2])
  }
```

```{r eval=TRUE, echo=FALSE}
b0.hat <- read_rds("../../rds/poweranalysis2_b0hat.rds")
b1.hat <- read_rds("../../rds/poweranalysis2_b1hat.rds")
b0.hat.calib<- read_rds( "../../rds/poweranalysis2_b0hatcalib.rds")
b1.hat.calib<- read_rds("../../rds/poweranalysis2_b0hatcalib.rds")
```

```{r}
hist(b0.hat, xlab = TeX("$\\hat{\\beta}_0$"),
     main = TeX("Histogram of the different $\\hat{\\beta}_0$ estimated in the synthetic datasets"))
hist(b1.hat, xlab = TeX("$\\hat{\\beta}_1$"),
     main = TeX("Histogram of the different $\\hat{\\beta}_1$ estimated in the synthetic datasets"))

paste0("In the 1000 simulated scenarios, the confidence interval included the true beta_0 ",
       100*mean(b0.hat.calib), "% of the times.", sep = "")
paste0("In the 1000 simulated scenarios, the confidence interval included the true beta_1 ",
       100*mean(b1.hat.calib), "% of the times.", sep = "")
```


## Power analysis demonstration  

We'll simulate the design of an experiment. 
Let's assume the following:  

- 3 $\times$ 2 factorial treatment structure, with temperature at 250, 400, and 500 F, and 2 muffin recipes. 

- Treatment means: $\mu_{11}=2$, $\mu_{21}=2.4$, $\mu_{31}=2.6$, $\mu_{12}=2.4$, $\mu_{22}=2.8$, $\mu_{32}=3$. 
- Which means: $\mu_{1\cdot} - \mu_{2\cdot} = -0.4$, $\mu_{1\cdot} - \mu_{2\cdot} = -0.4$.

**Consider these competing designs:** 

- CRD 
- split-plot design
- split-plot design with subsampling 
 
All examples below generate 100 designs with the means described above, the designs with 2 or 3 repetitions, and the corresponding variances for each design (e.g., $\sigma^2_\varepsilon$, $\sigma^2_{whole plot}$, etc.).   

### CRD - 2 reps 

```{r warning=FALSE, message=FALSE}
df_crd <- expand.grid(oven_temp = factor(c(250, 400, 500)),
                      recipe = c("B", "C"), 
                      rep = 1:2)
df_crd <- df_crd %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26


n_sims <- 100
p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)

n <- nrow(df_crd)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  df_temp <- df_crd %>% mutate(y = mu + rnorm(n, 0, sigma_epsilon))
  
  m <- lm(y ~ oven_temp * recipe, data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_1.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_2.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2,
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a CRD with 2 reps"))
hist(p_mu1._mu2., 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a CRD with 2 reps"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference in recipe
mean(p_mu1._mu2.<0.05) # power for .4 difference in temp
```

### CRD - 3 reps 

```{r warning=FALSE, message=FALSE}
df_crd <- expand.grid(oven_temp = factor(c(250, 400, 500)),
                      recipe = c("B", "C"), 
                      rep = 1:3)
df_crd <- df_crd %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26


p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)

n <- nrow(df_crd)

set.seed(42)
```


```{r warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}
for (i in 1:n_sims){
  df_temp <- df_crd %>% mutate(y = mu + rnorm(n, 0, sigma_epsilon))
  
  m <- lm(y ~ oven_temp * recipe, data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_3.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_4.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2, 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a CRD with 3 reps"))

hist(p_mu1._mu2., 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a CRD with 3 reps"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference in recipe
mean(p_mu1._mu2.<0.05) # power for .4 difference in temperature
```

### Split-plot - 2 reps 

```{r warning=FALSE, message=FALSE}
df_splitplot <- 
  expand.grid(oven_temp = factor(c(250, 400, 500)),
              recipe = c("B", "C"),
              rep = 1:2) %>% 
  mutate(wp_id = as.numeric(as.factor(paste(oven_temp, rep))))

df_splitplot <- df_splitplot %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26
sigma_oven <- .12
n <- nrow(df_splitplot)
n_wp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep)

p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)
```


```{r warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}
set.seed(42)

for (i in 1:n_sims){
  oven_re <- rnorm(n_wp, 0, sigma_oven)
  df_temp <- df_splitplot %>% mutate(y = mu + oven_re[wp_id] + rnorm(n, 0, sigma_epsilon) )
  
  m <- lmer(y ~ oven_temp * recipe + (1|oven_temp:rep), data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_5.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_6.rds")
```

```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2,
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a split-plot design with 2 reps"))
hist(p_mu1._mu2., 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a split-plot design with 2 reps"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference at the split plot
mean(p_mu1._mu2.<0.05) # power for .4 difference at the whole plot
```


### Split-plot - 3 reps 

```{r warning=FALSE, message=FALSE}
df_splitplot <- 
  expand.grid(oven_temp = factor(c(250, 400, 500)),
              recipe = c("B", "C"),
              rep = 1:3) %>% 
  mutate(wp_id = as.numeric(as.factor(paste(oven_temp, rep))))

df_splitplot <- df_splitplot %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26
sigma_oven <- .12
n <- nrow(df_splitplot)
n_wp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep)

p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  oven_re <- rnorm(n_wp, 0, sigma_oven)
  df_temp <- df_splitplot %>% mutate(y = mu + oven_re[wp_id] + rnorm(n, 0, sigma_epsilon) )
  
  m <- lmer(y ~ oven_temp * recipe + (1|oven_temp:rep), data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_7.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_8.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2, 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a split-plot design with 3 reps"))
hist(p_mu1._mu2.,  xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a split-plot design with 3 reps"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference at the split plot
mean(p_mu1._mu2.<0.05) # power for .4 difference at the whole plot
```

### Split-plot - 2 reps + subsampling

```{r warning=FALSE, message=FALSE}
df_splitplot_subsample <- 
  expand.grid(oven_temp = factor(c(250, 400, 500)),
              recipe = c("B", "C"),
              rep = 1:2, 
              subsample = 1:3) %>% 
  mutate(wp_id = as.numeric(as.factor(paste(rep, oven_temp))), 
         muffin_id = as.numeric(as.factor(paste(rep, oven_temp, recipe))))

df_splitplot_subsample <- df_splitplot_subsample %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26
sigma_oven <- .12
sigma_muffins <- .1

n <- nrow(df_splitplot)
n_wp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep)
n_sp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep) * n_distinct(df_splitplot$recipe)

p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)
```


```{r warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}
set.seed(42)

for (i in 1:n_sims){
  oven_re <- rnorm(n_wp, 0, sigma_oven)
  muffins_re <- rnorm(n_sp, 0, sigma_muffins)
  df_temp <- df_splitplot_subsample %>% 
    mutate(y = mu + oven_re[wp_id] + muffins_re[muffin_id] +
             rnorm(n, 0, sigma_epsilon) )
  
  m <- lmer(y ~ oven_temp * recipe + (1|oven_temp:rep/recipe), data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_9.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_10.rds")
```

```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2, 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a split-plot design with 2 reps and subsampling"))
hist(p_mu1._mu2.,  
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a split-plot design with 2 reps and subsampling"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference at the split plot
mean(p_mu1._mu2.<0.05) # power for .4 difference at the whole plot
```

### Split-plot - 3 reps + subsampling

```{r warning=FALSE, message=FALSE}
df_splitplot_subsample <- 
  expand.grid(oven_temp = factor(c(250, 400, 500)),
              recipe = c("B", "C"),
              rep = 1:4, 
              subsample = 1:3) %>% 
  mutate(wp_id = as.numeric(as.factor(paste(rep, oven_temp))), 
         muffin_id = as.numeric(as.factor(paste(rep, oven_temp, recipe))))

df_splitplot_subsample <- df_splitplot_subsample %>%
  mutate(mu = case_when(oven_temp == "250" & recipe == "B" ~ 2, 
                        oven_temp == "400" & recipe == "B" ~ 2.4, 
                        oven_temp == "500" & recipe == "B" ~ 2.6, 
                        oven_temp == "250" & recipe == "C" ~ 2.4, 
                        oven_temp == "400" & recipe == "C" ~ 2.8, 
                        oven_temp == "500" & recipe == "C" ~ 3))

sigma_epsilon <- 0.26
sigma_oven <- .12
sigma_muffins <- .1

n <- nrow(df_splitplot)
n_wp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep)
n_sp <- n_distinct(df_splitplot$oven_temp) * n_distinct(df_splitplot$rep) * n_distinct(df_splitplot$recipe)

p_mu.1_mu.2 <- numeric(n_sims)
p_mu1._mu2. <- numeric(n_sims)
```


```{r warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}
set.seed(3)

for (i in 1:n_sims){
  oven_re <- rnorm(n_wp, 0, sigma_oven)
  muffins_re <- rnorm(n_sp, 0, sigma_muffins)
  df_temp <- df_splitplot_subsample %>% 
    mutate(y = mu + oven_re[wp_id] + muffins_re[muffin_id] +
             rnorm(n, 0, sigma_epsilon) )
  
  m <- lmer(y ~ oven_temp * recipe + (1|oven_temp:rep/recipe), data = df_temp)
  p_mu.1_mu.2[i] <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1)))$contrasts)$p.value
  p_mu1._mu2.[i] <- as.data.frame(emmeans(m, ~oven_temp, contr = list(c(1, -1, 0)))$contrasts)$p.value
}
```

```{r eval=TRUE, echo=FALSE}
p_mu.1_mu.2 <- read_rds("../../rds/recipe_diff_11.rds")
p_mu1._mu2. <- read_rds("../../rds/recipe_diff_12.rds")
```

```{r warning=FALSE, message=FALSE}
hist(p_mu.1_mu.2, 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{\\cdot 1} - \\hat{\\mu}_{\\cdot 2}$ under a split-plot design with 3 reps and subsampling"))
hist(p_mu1._mu2., 
     xlim = c(0, 1),
     xlab = TeX("$\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1 \\cdot} - \\hat{\\mu}_{2 \\cdot}$ under a split-plot design with 3 reps and subsampling"))
```

```{r}
mean(p_mu.1_mu.2<0.05) # power for .4 difference at the split plot
mean(p_mu1._mu2.<0.05) # power for .4 difference at the whole plot
```



