# Planning a multi-location design
July 17th, 2025  

## Announcements

- HW 4 due tomorrow 
- Project due next Wednesday (July 23) for peer review. 

## Planning a multi-environment trial 

- Objectives: 
  - Cover more variability in the target environment 
  - Increase number of repetitions  
- Different approaches to picking the multiple locations: 
  - Carefully select multiple environments 
  - Randomly*(ish)* select multiple environments within the target environment 
  - How do both affect the response? 
  
### On the importance of sample size 

- "The truth wears off" by Jonah Lehrer [[link](https://www.newyorker.com/magazine/2010/12/13/the-truth-wears-off)] | [[alt. link](https://sites.ualberta.ca/~ahamann/teaching/renr480/reading/Lehrer_2010_The_truth_wears_off.pdf)] 
- How much are we relying on our results? Which results? 
- Sample size, design, etc. are important and affect conclusions! 


### What do the results for different multi-environment trials look like?  

Below is a simulation comparing 3 different scenarios of multi-environment trials with different number of environments tested. 
We assume that, within each environment, the trial is an RCBD with three repetitions. 
*[Context: this is perhaps the most common design in agronomy.]* 

We assume the following data generating process: 
$$y_{ijk} = \mu_i + b_{j(k)} + u_k +\varepsilon_{ijk}, \\
b_{j(k)} \sim N(0, \sigma_{b}^2), \\
u_k \sim N(0, \sigma_{u}^2), \\
\varepsilon_{ijk} \sim N(0, \sigma_{\varepsilon}^2),$$
where:

- $\mu_i$ is the treatment mean for the $i$th treatment ($i = 1,2, ..., 30$), 
- $b_{j(k)}$ is the effect of the $j$th block in the $k$th environment ($j = 1,2,3$), 
- $u_k$ is the effect of the $k$th environment ($k = 1,2,..., E$), and 
- $\varepsilon_{ijk}$ is the residual for the observation corresponding to the $i$th treatment, $j$th block in the $k$th environment. 

For this simulation, we assume that all $b_{j(k)}$, $u_k$ and $\varepsilon_{ijk}$ arise from independent normal distributions, with variances $\sigma_{b}^2 = 3$, $\sigma_{u}^2 = 35$, $\sigma_{\varepsilon}^2 = 10$.

Each simulation consists of 3 steps: 

1. Set "true" states (i.e., set $\mu_i$). We assume the same "ground truth" for all cases. To demonstrate the properties of the designs, we will assume some treatments have no difference (i.e., $\mu_i = \mu_{i'}$) and others are different (i.e., $\mu_i \neq \mu_{i'}$). *Note: this step does not depend on sample size.* 
2. Draw random samples from $b_{j(k)}$, $u_k$ and $\varepsilon_{ijk}$ to simulate observed values $y_{ijk}$. *Note: this step depends on sample size.*
3. Estimate the $\mu_i$s and test the type I error rate (count of times when $p<0.05$ when the truth was actually $\mu_i = \mu_{i'}$ -- reject $H_0$ when it was true), and the statistical power (count of times when $p<0.05$ when the truth was $\mu_i \ne \mu_{i'}$ -- reject $H_0$ when it was false). 


**R packages required for this simulation** 
```{r message=FALSE, warning=FALSE}
library(tidyverse) # data wrangling & data viz  
library(lme4) # model fitting
library(emmeans) # marginal means 
```

**Simulate 100 hypothetical experiments for 3 scenarios:** 

- **10 environments, 3 blocks each**

<details>
<summary>Click to show code for the simulation</summary>

```{r message=FALSE, warning=FALSE, eval=FALSE, echo=TRUE}
sigma_env.truth <- 35 
sigma_block.truth <- 3 
sigma.truth <- 10 

trts <- paste("t", 11:40, sep = "")
n_envs <- 10
n_rep <- 3
n_sims <- 250

df_me <- expand.grid(trt = trts,
                     rep = 1:n_rep, 
                     environm = paste("e", 11:(10+n_envs), sep = ""))

pval_0diff <- numeric(n_sims)
pval_20diff <- numeric(n_sims)

for (i in 1:n_sims) {
  env_re <- rnorm(n_envs, 0, sigma_env.truth)
  block_re <- rnorm(n_envs*n_rep, 0, sigma_block.truth)
  
  df_envs <- data.frame(environm = paste("e", 11:(10+n_envs), sep = ""), 
                        env_re)
  df_blocks <- expand.grid(environm = paste("e", 11:(10+n_envs), sep = ""),
                           rep = 1:n_rep) %>% 
    mutate(block_re = block_re)
  
  df_me <- df_me %>%
    mutate(mu_t = case_when(trt == "t11" ~ 200, 
                            trt == "t12" ~ 200, 
                            trt == "t13" ~ 180, 
                            .default = 220)) %>% 
    left_join(df_envs) %>% 
    left_join(df_blocks) %>% 
    mutate(e = rnorm(nrow(df_me), 0, sigma.truth), 
           y = mu_t + env_re + block_re + e)
  
  
  m <- lmer(y ~ trt + (1|environm/rep), data = df_me)
  pval_0diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, -1, rep(0, 28))))$contr)$p.value
  pval_20diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, 0, 0, -1, rep(0, 26))))$contr)$p.value
} 
```

</details>

```{r eval=TRUE, echo=FALSE}
pval_0diff <- read_rds("../../rds/pval_0diff_sim1.rds")
pval_20diff <- read_rds("../../rds/pval_20diff_sim1.rds")
```

```{r eval=TRUE, echo=FALSE}
hist(pval_0diff, main = "P-values testing a difference that, in reality, is zero. E = 10", 
     breaks = seq(0, 1, by = 0.05))
```

```{r eval=TRUE, echo=FALSE}
hist(pval_20diff, main = "P-values testing a difference that, in reality, is 20. E = 10", 
     breaks = seq(0, 1, by = 0.05))
```


```{r}
# type I error rate
mean(pval_0diff<0.05)
# statistical power
mean(pval_20diff<0.05)
```

- **2 environments, 3 blocks each**

<details>
<summary>Click to show code for the simulation</summary>

```{r message=FALSE, warning=FALSE, eval=FALSE, echo=TRUE}
sigma_env.truth <- 35 
sigma_block.truth <- 3 
sigma.truth <- 10 

trts <- paste("t", 11:40, sep = "")
n_envs <- 2
n_rep <- 3
n_sims <- 100

df_me <- expand.grid(trt = trts,
                     rep = 1:n_rep, 
                     environm = paste("e", 11:(10+n_envs), sep = ""))

pval_0diff <- numeric(n_sims)
pval_20diff <- numeric(n_sims)

set.seed(3)
for (i in 1:n_sims) {
  env_re <- rnorm(n_envs, 0, sigma_env.truth)
  block_re <- rnorm(n_envs*n_rep, 0, sigma_block.truth)
  
  df_envs <- data.frame(environm = paste("e", 11:(10+n_envs), sep = ""), 
                        env_re)
  df_blocks <- expand.grid(environm = paste("e", 11:(10+n_envs), sep = ""),
                           rep = 1:n_rep) %>% 
    mutate(block_re = block_re)
  
  df_me <- df_me %>%
    mutate(mu_t = case_when(trt == "t11" ~ 200, 
                            trt == "t12" ~ 200, 
                            trt == "t13" ~ 180, 
                            .default = 220)) %>% 
    left_join(df_envs) %>% 
    left_join(df_blocks) %>% 
    mutate(e = rnorm(nrow(df_me), 0, sigma.truth), 
           y = mu_t + env_re + block_re + e)
  
  
  m <- lmer(y ~ trt + (1|environm/rep), data = df_me)
  pval_0diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, -1, rep(0, 28))))$contr)$p.value
  pval_20diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, 0, 0, -1, rep(0, 26))))$contr)$p.value
}
```

</details>


```{r eval=TRUE, echo=FALSE}
pval_0diff <- read_rds("../../rds/pval_0diff_sim2.rds")
pval_20diff <- read_rds("../../rds/pval_20diff_sim2.rds")
```

```{r eval=TRUE, echo=FALSE}
hist(pval_0diff, main = "P-values testing a difference that, in reality, is zero. E = 2", 
     breaks = seq(0, 1, by = 0.05))
```

```{r eval=TRUE, echo=FALSE}
hist(pval_20diff, main = "P-values testing a difference that, in reality, is 20. E = 2", 
     breaks = seq(0, 1, by = 0.05))
```

```{r}
# type I error rate
mean(pval_0diff<0.05)
# statistical power
mean(pval_20diff<0.05)
```

- **1 environment, 3 blocks each**

<details>
<summary>Click to show code for the simulation</summary>

```{r message=FALSE, warning=FALSE, eval=FALSE, echo=TRUE}
sigma_env.truth <- 35 
sigma_block.truth <- 3 
sigma.truth <- 10 

trts <- paste("t", 11:40, sep = "")
n_envs <- 1
n_rep <- 3
n_sims <- 100

df_me <- expand.grid(trt = trts,
                     rep = 1:n_rep, 
                     environm = paste("e", 11:(10+n_envs), sep = ""))

pval_0diff <- numeric(n_sims)
pval_20diff <- numeric(n_sims)

set.seed(3)
for (i in 1:n_sims) {
  env_re <- rnorm(n_envs, 0, sigma_env.truth)
  block_re <- rnorm(n_envs*n_rep, 0, sigma_block.truth)
  
  df_envs <- data.frame(environm = paste("e", 11:(10+n_envs), sep = ""), 
                        env_re)
  df_blocks <- expand.grid(environm = paste("e", 11:(10+n_envs), sep = ""),
                           rep = 1:n_rep) %>% 
    mutate(block_re = block_re)
  
  df_me <- df_me %>%
    mutate(mu_t = case_when(trt == "t11" ~ 200, 
                            trt == "t12" ~ 200, 
                            trt == "t13" ~ 180, 
                            .default = 220)) %>% 
    left_join(df_envs) %>% 
    left_join(df_blocks) %>% 
    mutate(e = rnorm(nrow(df_me), 0, sigma.truth), 
           y = mu_t + env_re + block_re + e)
  
  
  m <- lmer(y ~ trt + (1|rep), data = df_me)
  pval_0diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, -1, rep(0, 28))))$contr)$p.value
  pval_20diff[i] <- as.data.frame(emmeans(m, ~ trt, contr = list(c(1, 0, 0, -1, rep(0, 26))))$contr)$p.value
}
```

</details>

```{r eval=TRUE, echo=FALSE}
pval_0diff <- read_rds("../../rds/pval_0diff_sim3.rds")
pval_20diff <- read_rds("../../rds/pval_20diff_sim3.rds")
```

```{r eval=TRUE, echo=FALSE}
hist(pval_0diff, main = "P-values testing a difference that, in reality, is zero. E = 1", 
     breaks = seq(0, 1, by = 0.05))
```

```{r eval=TRUE, echo=FALSE}
hist(pval_20diff, main = "P-values testing a difference that, in reality, is 20. E = 1", 
     breaks = seq(0, 1, by = 0.05))
```

```{r}
# type I error rate
mean(pval_0diff<0.05)
# statistical power
mean(pval_20diff<0.05)
```



