# Statistical inference and experiment designs 

JUne 23rd, 2026

## Announcements 

- Homework is due today at midnight. 
- Tomorrow will be online. 

## Why do we care about standard errors and degrees of freedom? 

### The hypothesis test explanation  

Typically (for a two-tailed test): 

1. Obtain t test statistic $t^\star = \frac{\hat\theta}{se(\hat\theta)}$.
2. Obtain $t_c = t_{df, \ \alpha/2}$. 
3. Obtain $P(t^\star>t_c)$. 
4. If $P(t^\star>t_c) > \alpha$, reject $H_0$.  

### The signal & noise explanation 

- Each point estimate should be associated to some measure of uncertainty 
- How little uncertainty do you need to be convinced? 
- What is a 100% confidence interval (CI)? Why is it less useful than a 95% CI?
- A 95% CI tells us the range of values where we would not reject the null ([Murtaugh, 2013](https://esajournals.onlinelibrary.wiley.com/doi/10.1890/13-0590.1)). 


### Applied case I -- CRD

A group of scientists ran an experiment to answer whether swine diets can affect the Serum haptoglobin concentration. The treatment structure was a $6 \times 2$ factorial with Diet (6 levels) and Moment (2 levels) as the treamtment factors. The design structure was a CRD with 16 repetitions.  
The scientists measured Serum haptoglobin (mg/dL) as the response variable and wish to maintain this metric as high as possible.  

```{r}
url <- "https://raw.githubusercontent.com/stat720/summer2026/refs/heads/main/data/blood_study_pigs.csv"
df <- read.csv(url) |> 
  mutate(Trt = as.factor(Trt), Day = as.factor(Day))
```

Because this is a CRD, the model is 

$$y_{ijk} = \mu_0 + T_i + D_j +(T\times D)_{ij} + \varepsilon_{ijk}, \\  \varepsilon_{ijk} \sim N(0, \sigma^2).$$

```{r}
m <- lm(Serum_haptoglobin_mg.dL ~ Trt*Day, data = df)

performance::check_model(m,  check = c("normality", "homogeneity"))

car::Anova(m, type =2)
```


### Biological significance (and others) matter as much as statistical significance  


- [ASA's statement on p-values](https://www.tandfonline.com/doi/full/10.1080/00031305.2016.1154108) 

```{r}
means <- emmeans(m, ~ Trt , at = list(Day = factor(38)))
plot(means)
```

How is the story behind ANOVA and the marginal means so different?


### Applied case II -- RCBD

- Field experiment studying the effect of potassium on corn yield.
- One treatment factor: Potassium fertilizer (one-way treatment structure).
- Randomized Complete Block Design with 4 repetitions (design structure).


```{r}
url <- "https://raw.githubusercontent.com/stat720/summer2026/refs/heads/main/data/cochrancox_kfert.csv"

df2 <- read.csv(url) |> 
  transmute(K2O_lbac = as.factor(K2O_lbac), rep = as.factor(rep), yield)
```

Because this is an RCBD, the model is 

$$y_{ij} = \mu_0 + T_i + b_j + \varepsilon_{ij}, \\  \varepsilon_{ij} \sim N(0, \sigma^2).$$

**Blocks: fixed of random?** 

- The assumptions behind $b_j$ vary: 
- Fundamentals of mixed models 
  - Assumption behind blocks as fixed effects: there is a 'true' block effect out there. 
  - Assumption behind blocks as random effects: 
- Confidence intervals of the means differ depending on the model:  
  - Blocks as fixed: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$
  - Blocks as random: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$ 
- Confidence intervals of the means differences don't differ depending on the model:  
  - Blocks as fixed/as random: $\hat\mu \pm t \cdot \sqrt{\frac{2 \sigma_{\varepsilon}^2 }{b}}$  


Blocks as fixed: 
```{r}
m_fixed <- lm(yield ~ K2O_lbac + rep, data = df2)
emmeans(m_fixed, ~ K2O_lbac, contr = list(c(1, -1, 0, 0, 0)))
```

Blocks as random: 
```{r}
m_random <- lmer(yield ~ K2O_lbac + (1|rep), data = df2)
emmeans(m_random, ~ K2O_lbac, contr = list(c(1, -1, 0, 0, 0)))
```

Things to highlight: 

- Under an RCBD, se(mean) $\neq$ se(mean difference). 
- Standard errors of the means differ depending on the model:  
  - Blocks as fixed: $\sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$
  - Blocks as random: $\sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$ 
- Standard errors of the means differences don't differ depending on the model:  
  - Blocks as fixed/as random: $\sqrt{\frac{2 \sigma_{\varepsilon}^2 }{b}}$  



## Applied case III -- split-plot design 

- Designed experiment of barley (70 levels) with fungicide (2 levels) treatments. 
- Two treatment factors: barley (70 levels) and fungicide (2 levels). Two-way factorial treatment structure.
- Split-plot design in an RCBD: Fungicide in the whole plot and genotype in the subplot.  


```{r}
agridat::durban.splitplot
df3 <- durban.splitplot
```


Because this is a split-plot design, the model is 

$$y_{ij} = \mu_0 + F_i + G_j + (F\times G)_{ij} + b_k + wp_{i(k)} + \varepsilon_{ij}, \\
wp_{i(k)}, \sim N(0, \sigma_{wp}^2) \\ \varepsilon_{ij} \sim N(0, \sigma^2_\varepsilon).$$

- Like before, we can choose either fixed or random for $b_k$. 
- But now, the se(mean) depend on the level we are looking.  

For differences at the whole-plot level (fung): $se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon + g \ \sigma^2_{wp})}{b \cdot g}}$. 
For differences at the split-plot level (gen): $se(\hat\mu-\mu') = \sqrt{\frac{2(\sigma^2_\varepsilon)}{b \cdot f}}$. 

```{r message=F}
m3 <- lmer(yield ~ gen*fung + (1|block/fung), data = df3)

emmeans(m3, ~fung, contr = list(c(1, -1)))$contr
emmeans(m3, ~gen, contr = list(c(1, -1, rep(0, 68))))$contr
```

```{r}
vc <- VarCorr(m3)
sigma2_b <- vc$block[1]
sigma2_wp <- vc$`fung:block`[1]
sigma2_epsilon <- sigma(m3)^2
n_block <- n_distinct(df3$block)
n_fung <- n_distinct(df3$fung)
n_gen <- n_distinct(df3$gen)

#se fung difference 
sqrt( 2*(sigma2_epsilon + (n_gen*sigma2_wp)) / (n_block *n_gen) )

#se gen difference 
sqrt( 2*(sigma2_epsilon ) / (n_block *n_fung) )
```

## Tomorrow  

- Online 
- Types of SS (I, II, III) and the inference 

