# Review: Applied split-plot 
July 2nd, 2025  

## Announcements

- Planning to miss >2 classes in July? [survey](https://forms.gle/TwhpwMD9i7ybawSa9) 
- Watch last week's classes (especially days 3+4) 
- Homework 2 due today. 
- Homework 3 is posted and due next Friday (July 11). 

## Background  

Yesterday, we designed an experiment with an RCBD to figure out the best temperature to bake the muffins. Check out the original recipe [here](https://www.ouichefnetwork.com/2016/06/a-life-changing-banana-nut-muffin.html).

Yesterday we agreed on the following model: 

$$y_{ik} = \mu + T_i + b_k + \varepsilon_{ik},$$

$$b_k \sim N(0, \sigma_b^2), \\ \varepsilon_{ik} \sim N(0, \sigma_{\varepsilon}^2),$$

and design: 
```{r echo=FALSE, message=FALSE, warning=FALSE}
t_design <- data.frame(Source = c("Day (block)", "", "Error(oven)", "Total"),
                       df = c("r-1 = 3-1 = 2", "-", "(t-1)r = (3-1)*3 = 6", "N-1 = 8"))

t_trt <- data.frame(Source = c(" ", "Temperature", "Parallels", "Total"),
                    df = c("-", "t-1 = 3-1 = 2", "N-t = 9-3 = 6", "N-1 = 8"))

t_rcbd <- data.frame(Source = c("Day", "Temperature", "Error(oven)", "Total"),
                    df = c("r-1 = 3-1 = 2", "t-1 = 3-1 = 2", "(t-1)r - (t-1) = 2*3 - 2 = 4", "N-1 = 8"))

knitr::kables(
  list(
    knitr::kable(t_design, caption = "Design Sources of Variability"),
    knitr::kable(t_trt, caption = "Treatment Sources of Variability"),
    knitr::kable(t_rcbd, caption = "Combined Table of the Sources of Variability")
  ),
  caption = ''
)
```

**Dry ingredients:**  

- 1 1/2 cups all-purpose flour (7 1/2 ounces)
- 1 1/2 teaspoons baking powder 
- 1/4 teaspoon baking soda 
- 1/2 teaspoon salt 
- 1/4 teaspoon ground cinnamon 
- 1/4 teaspoon ground nutmeg
- 2/3 cup finely chopped pecans, toasted

**Wet ingredients:**  

- 1/2 cup light brown sugar (3 3/4 ounces)
- 1/4 cup granulated sugar (2 ounces)
- 2 large eggs
- 1 stick unsalted butter, browned and cooled (4 ounces)
- 1 1/2 cups mashed bananas, from very ripe bananas that have been peeled and mashed well with a fork (12 3/4 ounces or 361 gr.)
- 2 tablespoons full-fat sour cream
- 1 teaspoon pure vanilla extract 

### Research question

What is the best temperature to bake the muffins? 

- 300 °F
- 400 °F
- 500 °F

How much banana? 

- 1 1/2 cups (12.75 oz, or 361 gr.)
- 2 cups (17 oz, or 482 gr.) 

## What is the best design? 

### Option A: another RCBD  

Corresponding model:


$$y_{ik} = \mu + T_i + B_j + (TB)_{ij} + b_k + \varepsilon_{ik},$$

$$b_k \sim N(0, \sigma_b^2), \\ \varepsilon_{ik} \sim N(0, \sigma_{\varepsilon}^2).$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
t_design <- data.frame(Source = c("Day (block)", "", "", "", "Error(oven)", "Total"),
                       df = c("r-1 = 3-1 = 2", "-", "-", "-", "(tb-1)*r = (6-1)*3 = 15", "N-1 = 17"))

t_trt <- data.frame(Source = c(" ", "Temperature", "Banana", "TxB", "Parallels", "Total"),
                    df = c("-", "t-1 = 3-1 = 2", "b-1 = 2-1 = 1", "(t-1)*(b-1) = 2", "N-tb = 18 - (3*2) = 12", "N-1 = 17"))

t_rcbd <- data.frame(Source = c("Day", "Temperature", "Banana", "TxB", "Error(oven)", "Total"),
                    df = c("r-1 = 3-1 = 2", "t-1 = 3-1 = 2", "b-1 = 2-1 = 1", "(t-1)*(b-1) = 2",
                           "(tb-1)*r - (t-1) - (b-1) - (t-1)*(b-1) = 15 - 2 -1 -2 = 10", "N-1 = 17"))

knitr::kables(
  list(
    knitr::kable(t_design, caption = "Design Sources of Variability"),
    knitr::kable(t_trt, caption = "Treatment Sources of Variability"),
    knitr::kable(t_rcbd, caption = "Combined Table of the Sources of Variability")
  ),
  caption = ''
)
```

### Option B: split-plot design

In the model:

- Cooking days are blocks. 
- Oven are "mini-blocks" for banana recipe. 

Statistical model:

$$y_{ik} = \mu + T_i + B_j + (TB)_{ij} + b_k + w_{i(k)} + \varepsilon_{ik},$$

$$b_k \sim N(0, \sigma_b^2), \\ w_{i(k)} \sim N(0, \sigma^2_w) , \\ \varepsilon_{ik} \sim N(0, \sigma_{\varepsilon}^2).$$


ANOVA table for this split-plot design 

```{r echo=FALSE, message=FALSE, warning=FALSE}
t_design <- data.frame(Source = c("Day (block)", "", "Error(oven)", 
                                  "", "", "Error(oven x day)", "Total"),
                       df = c("r-1 = 3-1 = 2", "-", "(t-1)*r = (3-1)*3 = 6",
                              "-", "-", "(b-1)* t * r = (2-1) * 3 * 3 = 9", "N-1 = 17"))

t_trt <- data.frame(Source = c(" ", "Temperature", "", 
                               "Banana", "TxB", "Parallels", "Total"),
                    df = c("-", "t-1 = 3-1 = 2", "-",
                           "b-1 = 2-1 = 1", "(t-1)*(b-1) = 2", "N-tb = 18 - (3*2) = 12", "N-1 = 17"))

t_rcbd <- data.frame(Source = c("Day", "Temperature", "Error(oven)",
                                "Banana", "TxB", "Error(oven x day)", "Total"),
                    df = c("r-1 = 3-1 = 2", "t-1 = 3-1 = 2", "(t-1)*r - (t-1)= 6 -2 = 4",
                           "b-1 = 2-1 = 1", "(t-1)*(b-1) = 2", 
                           "(b-1)* t * r - (b-1) - (t-1)*(b-1) = 9 - 1 -2 = 6", "N-1 = 17"))

knitr::kables(
  list(
    knitr::kable(t_design, caption = "Design or Topographical Sources of Variability"),
    knitr::kable(t_trt, caption = "Treatment Sources of Variability"),
    knitr::kable(t_rcbd, caption = "Combined Table of the Sources of Variability")
  ),
  caption = ''
)
```

## Treatment means and confidence intervals for the split-plot design  

The treatment mean for the $i$th temperature and $j$th banana level is $\mu_{ij} = \mu + T_i + B_j +(TB)_{ij}$. 
That mean won't change under different design structures. 
What may change is the confidence interval around the mean difference.  

- First, recall the formula for a CI: $\theta \pm t_{df, \frac{\alpha}{2}} \cdot se(\hat{\theta})$

```{r}
#get test t
qt(p = .975, df = 4) # df are df error(oven)
```

- For example, the CI for the differences between means for 300F and 400F $\mu_{1 \cdot} - \mu_{2 \cdot}$ is $(\mu_{1 \cdot} - \mu_{2 \cdot}) \pm 2.78 \cdot se(\widehat{\mu_{1 \cdot} - \mu_{2 \cdot}})$ 
- $se(\widehat{\mu_{1 \cdot} - \mu_{2 \cdot}}) = \sqrt{\frac{2 (\sigma^2_{\varepsilon} + b  \cdot \sigma^2_w)}{b \cdot r}}$

$$\mu_i \pm 2.78 \cdot \sqrt{\frac{2 (\sigma^2_{\varepsilon} + b \cdot \sigma^2_w)}{b \cdot r}}$$


```{r}
#get test t
qt(p = .975, df = 6) # df are df error(oven)
```

- The CI for the differences between means for normal and high banana $\mu_{\cdot 1} - \mu_{\cdot 2}$ is $(\mu_{\cdot 1} - \mu_{\cdot 2}) \pm 2.44 \cdot se(\widehat{\mu_{\cdot 1} - \mu_{\cdot 2}})$ 
- $se(\widehat{\mu_{\cdot 1} - \mu_{\cdot 2}}) = \sqrt{\frac{2 \sigma^2_{\varepsilon}}{t \cdot r}}$

$$\mu_i \pm 2.44 \cdot \sqrt{\frac{2 \sigma^2_{\varepsilon}}{r}}$$

## Tomorrow: 

- Measure heights of the muffins and analyze the data!  

