# Planning an experiment II 

July 2nd, 2026

```{r include=FALSE}
library(tidyverse)
library(emmeans)
library(latex2exp)
library(lme4)
```

## Announcements 

- Tomorrow is an observed holiday
- HW 3 is due on Tuesday 
- Projects are due July 12th. 

## Planning an experiment 

### Research question 

- What is the best recipe to prepare banana muffins? 

  - A: Baking soda only 
  - B: Baking soda + baking powder 
  - C: Baking powder only 

- What if it was a different combination?

### Experiment design 

What we know: 

- $\sigma^2_\varepsilon \approx 0.1$ 

**Analytic approach:**

```{r}
reps <- 5:40
sigma2_e <- .1
alpha <- .05

t_value <- qt(1-alpha/2, df = (reps-1)*2)
se_thetahat <- sqrt(2*sigma2_e / reps)

data.frame(reps, 
           CI_width = t_value*se_thetahat) |> 
  ggplot(aes(reps, CI_width))+
  geom_point()+
  coord_cartesian(ylim = c(0, .5))+
  scale_x_continuous(breaks = 1:40)+
  geom_hline(yintercept = .15)+
  labs(x = "# reps", y = "CI width")+
  theme_minimal()+
  theme(aspect.ratio = .4)
```


**Simulation-based approach:**

- About simulations [[link 1](https://royalsocietypublishing.org/rsta/article/384/2321/20240616/481849/Simulations-in-statistical-workflows)] [[link 2](https://onlinelibrary.wiley.com/doi/10.1002/sim.8086)]

**Consider these competing designs:** 

- RCBD with 10 reps with its corresponding statistical model, $y_{ij} = \mu + T_i + o_j + \varepsilon_{ij}$ 
- RCBD with 10 reps and subsampling with its corresponding statistical model, $y_{ij} = \mu + T_i + o_j + s_{i(j)} + \varepsilon_{ij}, s_{i(j)} \sim N(0, \sigma^2_o)$ 

All examples below generate 100 designs with the means described above.   

### RCBD with 10 reps 


The data generating model is **literally** $y_{ij} = \mu + T_i + o_j + \varepsilon_{ij}$, with $o_j \sim N(0, 0.05)$ and $\varepsilon_{ij} \sim N(0, 0.1)$.

```{r warning=FALSE, message=FALSE}
n_reps <- 10

df_rcbd <- expand.grid(recipe = c("A", "B", "C"), 
                      rep = 1:n_reps)
df_rcbd$rep_f <- as.factor(df_rcbd$rep)
df_rcbd <- df_rcbd %>%
  mutate(mu = case_when(recipe == "A" ~ 0, 
                        recipe == "B" ~ .15, 
                        recipe == "C" ~ .5))

sigma_epsilon <- sqrt(.1)
sigma_rep <- sqrt(.05)


n_sims <- 100
p_mu1_minus_mu2 <- numeric(n_sims)
p_mu1_minus_mu3 <- numeric(n_sims)
se_mu1_minus_mu2 <- numeric(n_sims)
se_mu1_minus_mu3 <- numeric(n_sims)

n <- nrow(df_rcbd)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  oven_eff <- rnorm(n_reps, 0, sigma_rep)
  df_temp <- df_rcbd |> mutate(y = mu + oven_eff[rep] + rnorm(n, 0, sigma_epsilon))
  
  m <- lm(y ~ recipe + rep_f, data = df_temp)
  means.temp <- as.data.frame(emmeans(m, ~recipe, contr = list(c(1, -1, 0), 
                                                               c(1, 0, -1)))$contrasts)
  
  p_mu1_minus_mu2[i] <- means.temp$p.value[1]
  p_mu1_minus_mu3[i] <- means.temp$p.value[2]
  se_mu1_minus_mu2[i] <- means.temp$SE[1]
  se_mu1_minus_mu3[i] <- means.temp$SE[2]}
```

```{r eval=TRUE, echo=FALSE}
# p_mu1_minus_mu2 |> write_rds("../../rds/p_mu1_minus_mu2.rds")
# p_mu1_minus_mu3 |> write_rds("../../rds/p_mu1_minus_mu3.rds")
# se_mu1_minus_mu2 |> write_rds("../../rds/se_mu1_minus_mu2.rds")
# se_mu1_minus_mu3 |> write_rds("../../rds/se_mu1_minus_mu3.rds")

p_mu1_minus_mu2 <- read_rds("../../rds/p_mu1_minus_mu2.rds")
p_mu1_minus_mu3 <- read_rds("../../rds/p_mu1_minus_mu3.rds")
se_mu1_minus_mu2 <- read_rds("../../rds/se_mu1_minus_mu2.rds")
se_mu1_minus_mu3 <- read_rds("../../rds/se_mu1_minus_mu3.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu1_minus_mu2,
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"))
hist(p_mu1_minus_mu3, 
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"))


hist(se_mu1_minus_mu2,
     xlim = c(0, .3),
     xlab = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"),
     main = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"))
hist(se_mu1_minus_mu3, 
     xlim = c(0, .3),
     xlab = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{3})$"),
     main = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{3})$"))
```

```{r}
mean(p_mu1_minus_mu2<0.05) # power for a .15 difference in recipe
mean(p_mu1_minus_mu3<0.05) # power for a .5 difference in temp
mean(se_mu1_minus_mu3) # se for a .5 difference in temp
```

### RCBD with 10 reps and 6 subsamples   

Here, the data generating model is **literally** $y_{ijk} = \mu + T_i + o_j + s_{i(j)} + \varepsilon_{ijk}$, with $o_j \sim N(0, 0.05)$, $s_{i(j)} \sim N(0, 0.01)$ and $\varepsilon_{ijk} \sim N(0, 0.1)$.


```{r warning=FALSE, message=FALSE}
n_reps <- 10
n_subsamps <- 6

df_rcbd <- expand.grid(recipe = c("A", "B", "C"), 
                      rep = 1:n_reps, 
                      subsamp = 1:n_subsamps)

df_rcbd <- df_rcbd %>%
  mutate(mu = case_when(recipe == "A" ~ 0, 
                        recipe == "B" ~ .15, 
                        recipe == "C" ~ .5), 
         rep_f = as.factor(rep), 
         ss = as.numeric(as.factor(paste(rep, recipe))))

sigma_epsilon <- sqrt(.3)
sigma_ss <- sqrt(.01)
sigma_rep <- sqrt(.05)

n_sims <- 100
p_mu1_minus_mu2 <- numeric(n_sims)
p_mu1_minus_mu3 <- numeric(n_sims)
se_mu1_minus_mu2 <- numeric(n_sims)
se_mu1_minus_mu3 <- numeric(n_sims)

n <- nrow(df_rcbd)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  oven_eff <- rnorm(n_reps, 0, sigma_rep)
  ss_eff <- rnorm(n_reps*n_subsamps, 0, sigma_ss)
  df_temp <- df_rcbd |> 
    mutate(y = mu + oven_eff[rep] + ss_eff[ss] + rnorm(n, 0, sigma_epsilon))
  
  m <- lmer(y ~ recipe + (1|rep/recipe), data = df_temp)
  
  means.temp <- as.data.frame(emmeans(m, ~recipe, 
                                      contr = list(c(1, -1, 0), 
                                                               c(1, 0, -1)))$contrasts)
  
  p_mu1_minus_mu2[i] <- means.temp$p.value[1]
  p_mu1_minus_mu3[i] <- means.temp$p.value[2]
  se_mu1_minus_mu2[i] <- means.temp$SE[1]
  se_mu1_minus_mu3[i] <- means.temp$SE[2]
  }

  means.temp
```

```{r eval=TRUE, echo=FALSE}
# p_mu1_minus_mu2 |> write_rds("../../rds/p_mu1_minus_mu2_subs.rds")
# p_mu1_minus_mu3 |> write_rds("../../rds/p_mu1_minus_mu3_subs.rds")
# se_mu1_minus_mu2 |> write_rds("../../rds/se_mu1_minus_mu2_subs.rds")
# se_mu1_minus_mu3 |> write_rds("../../rds/se_mu1_minus_mu3_subs.rds")

p_mu1_minus_mu2 <- read_rds("../../rds/p_mu1_minus_mu2_subs.rds")
p_mu1_minus_mu3 <- read_rds("../../rds/p_mu1_minus_mu3_subs.rds")
se_mu1_minus_mu2 <- read_rds("../../rds/se_mu1_minus_mu2_subs.rds")
se_mu1_minus_mu3 <- read_rds("../../rds/se_mu1_minus_mu3_subs.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu1_minus_mu2,
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"))
hist(p_mu1_minus_mu3, 
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"))


hist(se_mu1_minus_mu2,
     xlim = c(0, .3),
     xlab = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"),
     main = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"))
```

```{r}
mean(p_mu1_minus_mu2<0.05) # power for a .15 difference 
mean(p_mu1_minus_mu3<0.05) # power for a .5 difference  
mean(se_mu1_minus_mu3) # se for a .15 difference 
```


## Take-home messages 

- Two approaches to planning experiments and evaluating hypotheticals. 
- For a more complex example, check out [this example comparing RCBD and two different types of split-plot designs](https://stat720.github.io/summer2026/assignments/report_example_muffins.pdf).
- Read: Casler, M.D. (2015), Fundamentals of Experimental Design: Guidelines for Designing Successful Experiments. Agronomy Journal, 107: 692-705. [https://doi.org/10.2134/agronj2013.0114](https://doi.org/10.2134/agronj2013.0114)


