# Randomized complete block designs 

July 6th, 2026

## Announcements 

- HW 3 is due on Tomorrow 
- Tomorrow is online 
- Projects are due July 12th 

## RCBD review 

- Blocks are groups of approximately similar experimental units. 
- Data generated by blocked designs may be analyzed as $y_{ij} = \mu +\tau_i + b_j + \varepsilon_{ij}$, where $y_{ij}$ is the observed response for the $i$th treatment in the $j$th block, $\tau_i$ is the effect of the $i$th treatment ($i = 1, 2, ..., t$), $b_j$ is the effect of the $j$th block, $\varepsilon_{ij}$ is the residual of the $i$th treatment in the $j$th block. 
- Most RCBDs have $k = t$, where $k$ is the number of experimental units per block, and $t$ is the number of treatments. 

## Blocks - fixed or random?  

- The assumptions behind $b_j$ vary: 
- Fundamentals of mixed models 
  - Assumption behind blocks as fixed effects: there is a 'true' block effect out there. 
  - Assumption behind blocks as random effects: 
- Confidence intervals of the means differ depending on the model:  
  - Blocks as fixed: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 }{b}}$
  - Blocks as random: $\hat\mu \pm t \cdot \sqrt{\frac{\sigma_{\varepsilon}^2 + \sigma^2_{b}}{b}}$ 
- Confidence intervals of the means differences don't differ depending on the model:  
  - Blocks as fixed/as random: $\hat\mu \pm t \cdot \sqrt{\frac{2 \sigma_{\varepsilon}^2 }{b}}$  
- More on the differences between blocks as fixed vs random: [Dixon 2016](https://newprairiepress.org/cgi/viewcontent.cgi?article=1474&context=agstatconference). 

## Applied design and analysis of an RCBD   


From Thursday: RCBD with 10 reps and 6 subsamples   

The data generating model is now $y_{ijk} = \mu + T_i + o_j + s_{i(j)} + \varepsilon_{ijk}$, with $o_j \sim N(0, 0.15)$, $s_{i(j)} \sim N(0, 0.02)$ and $\varepsilon_{ijk} \sim N(0, 0.1)$. Note that $\sigma^2_{oven}$ is now large, because it now captures the variability between ovens and houses, and we expect that to be a bit larger. 

```{r include=FALSE}
library(tidyverse)
library(emmeans)
library(latex2exp)
library(lme4)
```

```{r warning=FALSE, message=FALSE}
n_reps <- 10
n_subsamps <- 6

df_rcbd <- expand.grid(recipe = c("A", "B", "C"), 
                      rep = 1:n_reps, 
                      subsamp = 1:n_subsamps)

df_rcbd <- df_rcbd %>%
  mutate(mu = case_when(recipe == "A" ~ 0, 
                        recipe == "B" ~ .15, 
                        recipe == "C" ~ .5), 
         rep_f = as.factor(rep), 
         ss = as.numeric(as.factor(paste(rep, recipe))))

sigma_epsilon <- sqrt(.3)
sigma_ss <- sqrt(.02)
sigma_rep <- sqrt(.15)

n_sims <- 100
p_mu1_minus_mu2 <- numeric(n_sims)
p_mu1_minus_mu3 <- numeric(n_sims)
se_mu1_minus_mu2 <- numeric(n_sims)
se_mu1_minus_mu3 <- numeric(n_sims)

n <- nrow(df_rcbd)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  oven_eff <- rnorm(n_reps, 0, sigma_rep)
  ss_eff <- rnorm(n_reps*n_subsamps, 0, sigma_ss)
  df_temp <- df_rcbd |> 
    mutate(y = mu + oven_eff[rep] + ss_eff[ss] + rnorm(n, 0, sigma_epsilon))
  
  m <- lmer(y ~ recipe + (1|rep/recipe), data = df_temp)
  
  means.temp <- as.data.frame(emmeans(m, ~recipe, 
                                      contr = list(c(1, -1, 0), 
                                                               c(1, 0, -1)))$contrasts)
  
  p_mu1_minus_mu2[i] <- means.temp$p.value[1]
  p_mu1_minus_mu3[i] <- means.temp$p.value[2]
  se_mu1_minus_mu2[i] <- means.temp$SE[1]
  se_mu1_minus_mu3[i] <- means.temp$SE[2]
  }
```

```{r eval=TRUE, echo=FALSE}
# p_mu1_minus_mu2 |> write_rds("../../rds/p_mu1_minus_mu2_subs_0706.rds")
# p_mu1_minus_mu3 |> write_rds("../../rds/p_mu1_minus_mu3_subs_0706.rds")
# se_mu1_minus_mu2 |> write_rds("../../rds/se_mu1_minus_mu2_subs_0706.rds")
# se_mu1_minus_mu3 |> write_rds("../../rds/se_mu1_minus_mu3_subs_0706.rds")

p_mu1_minus_mu2 <- read_rds("../../rds/p_mu1_minus_mu2_subs_0706.rds")
p_mu1_minus_mu3 <- read_rds("../../rds/p_mu1_minus_mu3_subs_0706.rds")
se_mu1_minus_mu2 <- read_rds("../../rds/se_mu1_minus_mu2_subs_0706.rds")
se_mu1_minus_mu3 <- read_rds("../../rds/se_mu1_minus_mu3_subs_0706.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu1_minus_mu2,
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"))
hist(p_mu1_minus_mu3, 
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"))


hist(se_mu1_minus_mu2,
     xlim = c(0, .3),
     xlab = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"),
     main = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"))
```

```{r}
mean(p_mu1_minus_mu2<0.05) # power for a .15 difference 
mean(p_mu1_minus_mu3<0.05) # power for a .5 difference  
mean(se_mu1_minus_mu3) # se for a .15 difference 
```


## Planning the experiment for Wednesday 

- Assign treatments and discuss the implications of the different treatment allocations (logistically) 
- Design a protocol

### Recipe

- 3/4 cup (94g) all-purpose flour
- 1 teaspoon chemical leavener (Trt A, B, or C):
  - Trt A: 1 teaspoon baking soda
  - Trt B: 1/2 teaspoon baking powder + 1/2 teaspoon baking soda
  - Trt C: 1 teaspoon baking powder
- 1/4 teaspoon salt
- 3/4 cup (173g) mashed bananas (about 2 medium or 1.5 large ripe bananas)
- 3 Tablespoons (43g) unsalted butter, melted
- 1/3 cup (68g) granulated sugar 
- 1 large egg, at room temperature

- Preheat oven to 425 F. 
- Whisk the flour, baking powder, baking soda, and salt together in a medium bowl. Set aside. 
- In a large bowl, mash the bananas. 
- On medium speed, beat or whisk in the melted butter, sugar, and egg. 
- Pour the dry ingredients into the wet ingredients, whisk until combined. 

- Spoon the batter into liners, filling them all the way to the top. 
- Bake for 5 minutes at 425°F, then, keeping the muffins in the oven, reduce the oven temperature to 350°F (177°C). 
- Bake for an additional 16–18 minutes or until a toothpick inserted in the center comes out clean. The total time these muffins take in the oven is about 21–23 minutes, give or take. 
- Allow the muffins to cool for 5 minutes in the muffin pan, then transfer to a wire cooling rack to continue cooling.

- [See original recipe](https://sallysbakingaddiction.com/banana-muffins/)

