# Analyzing the in-class experiment 

July 8th, 2026

## Announcements 

- Make sure you find a classmate to do peer review with 
- Please leave your feedback in [this form](https://forms.gle/YdiV4B3kFfNJtmkLA) if you want to have some influence on next week's topics. 

## Our designed experiment 

RCBD with 10 reps and 6 subsamples   

The data generating model is now $y_{ijk} = \mu + T_i + o_j + s_{i(j)} + \varepsilon_{ijk}$, with $o_j \sim N(0, 0.15)$, $s_{i(j)} \sim N(0, 0.02)$ and $\varepsilon_{ijk} \sim N(0, 0.1)$. Note that $\sigma^2_{oven}$ is now large, because it now captures the variability between ovens and houses, and we expect that to be a bit larger. 


**What can we expect?**

-> Simulate data with $j = 1, 2, \dots, 11$, and $k = 1, 2, \dots, 4$.


```{r include=FALSE}
library(tidyverse)
library(emmeans)
library(latex2exp)
library(lme4)
```

```{r warning=FALSE, message=FALSE}
n_reps <- 11
n_subsamps <- 4

df_rcbd <- expand.grid(recipe = c("A", "B", "C"), 
                      rep = 1:n_reps, 
                      subsamp = 1:n_subsamps)

df_rcbd <- df_rcbd %>%
  mutate(mu = case_when(recipe == "A" ~ 0, 
                        recipe == "B" ~ .15, 
                        recipe == "C" ~ .5), 
         rep_f = as.factor(rep), 
         ss = as.numeric(as.factor(paste(rep, recipe))))

sigma_epsilon <- sqrt(.3)
sigma_ss <- sqrt(.02)
sigma_rep <- sqrt(.15)

n_sims <- 100
p_mu1_minus_mu2 <- numeric(n_sims)
p_mu1_minus_mu3 <- numeric(n_sims)
se_mu1_minus_mu2 <- numeric(n_sims)
se_mu1_minus_mu3 <- numeric(n_sims)

n <- nrow(df_rcbd)
```


```{r warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
set.seed(42)

for (i in 1:n_sims){
  oven_eff <- rnorm(n_reps, 0, sigma_rep)
  ss_eff <- rnorm(n_reps*n_subsamps, 0, sigma_ss)
  df_temp <- df_rcbd |> 
    mutate(y = mu + oven_eff[rep] + ss_eff[ss] + rnorm(n, 0, sigma_epsilon))
  
  m <- lmer(y ~ recipe + (1|rep/recipe), data = df_temp)
  
  means.temp <- as.data.frame(emmeans(m, ~recipe, 
                                      contr = list(c(1, -1, 0), 
                                                               c(1, 0, -1)))$contrasts)
  
  p_mu1_minus_mu2[i] <- means.temp$p.value[1]
  p_mu1_minus_mu3[i] <- means.temp$p.value[2]
  se_mu1_minus_mu2[i] <- means.temp$SE[1]
  se_mu1_minus_mu3[i] <- means.temp$SE[2]
  }
```

```{r eval=TRUE, echo=FALSE}
# p_mu1_minus_mu2 |> write_rds("../../rds/p_mu1_minus_mu2_subs_0708.rds")
# p_mu1_minus_mu3 |> write_rds("../../rds/p_mu1_minus_mu3_subs_0708.rds")
# se_mu1_minus_mu2 |> write_rds("../../rds/se_mu1_minus_mu2_subs_0708.rds")
# se_mu1_minus_mu3 |> write_rds("../../rds/se_mu1_minus_mu3_subs_0708.rds")

p_mu1_minus_mu2 <- read_rds("../../rds/p_mu1_minus_mu2_subs_0708.rds")
p_mu1_minus_mu3 <- read_rds("../../rds/p_mu1_minus_mu3_subs_0708.rds")
se_mu1_minus_mu2 <- read_rds("../../rds/se_mu1_minus_mu2_subs_0708.rds")
se_mu1_minus_mu3 <- read_rds("../../rds/se_mu1_minus_mu3_subs_0708.rds")
```


```{r warning=FALSE, message=FALSE}
hist(p_mu1_minus_mu2,
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{2}$"))
hist(p_mu1_minus_mu3, 
     xlim = c(0, 1),
     xlab = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"),
     main = TeX("p-value for $\\hat{\\mu}_{1} - \\hat{\\mu}_{3}$"))


hist(se_mu1_minus_mu2,
     xlim = c(0, .3),
     xlab = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"),
     main = TeX("se($\\hat{\\mu}_{1} - \\hat{\\mu}_{2})$"))
```

```{r}
mean(p_mu1_minus_mu2<0.05) # power for a .15 difference 
mean(p_mu1_minus_mu3<0.05) # power for a .5 difference  
mean(se_mu1_minus_mu3) # se for a .15 difference 
```

## A few extra (anticipated) comments 

- Postblocking 
  - Include blocks (i.e., groups of similar EUs) that were identified after randomization 
  - Typically we need to know the source of the post-blocks 
  - This is a slightly controversial topic en DoE -- see [Ainsley et al. (1987)](https://www.jstor.org/stable/2531948?seq=1)
  - See [Tadese and Piepho (2023)](https://acsess.onlinelibrary.wiley.com/doi/full/10.1002/agj2.21450) 

- Modeling unbalanced data using mixed models 
  - See [Spilke et al. (2005)](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1439-037X.2004.00120.x)

## Tomorrow 

- Online 
- Work on your projects!


