# Optimal designs II 

## Announcements 

- Tomorrow: In-person Kahoot!

## Optimal designs 

### Case study: incomplete block designs 

- We can know the treatment structure + design structure and still not know everything about the experiment design! 

**For example:** 

- One-way treatment structure + incomplete block design 
- There are multiple options to arrange the treatments in the incomplete blocks, leading to different advantages/disadvantages 
  
- Case study: One-way treatment structure with 4 levels + incomplete block design with block size = 2.

**Balanced incomplete block designs**  

- $Var (T_i - T_{i'})$ is constant 
- Each pair of treatments occur together (i.e., in the same block) $\lambda$ times 
- Not all designs (i.e., combination of block size, #reps, and #treatments) can make this possible 

Assume: 

- $t=$ number of treatments, 
- $b=$ number of blocks, 
- $k=$ block size, 
- $r=$ number of repetitions,
- Thus, $N=kb$ (where $N$ is the number of total observations)
- To have balance (i.e., "each pair of treatments occur $\lambda$ times in the same block"):
  - $\lambda = \frac{r(k-1)}{t-1}$ 
  - BIBDs can be built with the `ibd` R package [[link](https://cran.r-project.org/web/packages/ibd/index.html)] 

  
This is easy for this case with $t=4, \ b=6, \ k =2, r =3$:

- $\lambda = \frac{3(2-1)}{4-1} = \frac{3}{3}=1$. 
- See below.

```{r echo=FALSE}
table1 <- data.frame(
  blocks = 1:6,
  A = rep(c("X", ""), each =3),
  B = c("X", "", "", "X", "X", ""),
  C = c("", "X", "", "X","",  "X"),
  D = c("", "", "X", "", "X","X")
)
table1 <- as.data.frame(t(table1))[-1,]
colnames(table1) <- paste("Block ", 1:6)

knitr::kable(table1, 
             caption = "Assignment of Four Treatments to Six Blocks of Two Experimental
Units. Balanced Incomplete Block Example. [Adapted from Table 4.19 in Milliken and Johnson]")
```

The $\mathbf{X}$ matrix above minimizes $(\mathbf{X}^T\mathbf{X})^{-1}$:

$$\mathbf{X} = \begin{bmatrix} 1 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
1 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
1 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
0 & 1 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & 1 & 0 & 0 & 0 & 0 & 0\\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 1
\end{bmatrix}$$

With that, we minimize the variance to test $\mu_A =\mu_B =\mu_C =\mu_D$.


**Unbalanced incomplete block designs**  

- $Var (T_i - T_{i'})$ is not constant 
- Each pair of treatments do not occur together (i.e., in the same block) the same number of times 
- Prioritize the comparisons of interest 

See below: D-optimal experiment to test if $\mu_A = \mu_B, \mu_A = \mu_C, \mu_A = \mu_D$.

```{r echo=FALSE}
table2 <- data.frame(
  blocks = 1:6,
  A = rep("X", 6),
  B = c("X", "", "", "X", "", ""),
  C = c("", "X", "", "",  "X", ""),
  D = c("", "", "X", "", "","X")
)
table2 <- as.data.frame(t(table2))[-1,]
colnames(table2) <- paste("Block ", 1:6)


knitr::kable(table2, 
             caption = "Assignment of Four Treatments to Six Blocks of Two Experimental
Units. Unbalanced Incomplete Block Example. [Adapted from Table 4.22 in Milliken and Johnson]")
```

**Disconnected unbalanced incomplete block designs**  

- See below. Neither Treatment A nor Treatment B occur in a block with either treatment C or D.

```{r echo=FALSE}
table3 <- data.frame(
  blocks = 1:6,
  A = c("X", "", "X", "", "X", ""),
  B = c("X", "", "X", "", "X", ""),
  C = c("", "X", "", "X", "",  "X"),
  D = c("", "X", "", "X", "",  "X")
)
table3 <- as.data.frame(t(table3))[-1,]
colnames(table3) <- paste("Block ", 1:6)


knitr::kable(table3, 
             caption = "Assignment of Four Treatments to Six Blocks of Two Experimental
Units. Disconnected Unbalanced Incomplete Block Example. [Adapted from Table 4.23 in Milliken and Johnson]")
```

