# More multilevel designs
July 16th, 2025  

## Announcements

- HW 4 due Friday 

## Review: statistical models  

$$y = \text{Treatment sources of variability} + \text{Design sources of variability} + \varepsilon$$


## Multi-location trials  

- Nature of multi-location trials 
- Objectives of a multi-location trial  
- How many years? 
- Exercise: find the experimental unit and the groups that are not independent. 

### Example  

Multi-environment trial of 64 corn hybrids in six counties in North Carolina. Each location had 3 replicates in in incomplete-block design. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(agridat)
library(lme4)
library(emmeans)

df_multienv <- agridat::besag.met

df_multienv %>% 
  ggplot(aes(col, row))+
  geom_tile(aes(fill = block))+
  facet_wrap(~county)

m_multienv <- lmer(yield ~ gen + (1|county/block), 
                    data = df_multienv)

VarCorr(m_multienv)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
t_met <- data.frame(Source = c("Location", 
                                "Block (Location)",
                                "Genotype",
                                "Error",
                                "Total"),
                    df = c("l - 1 = 5",
                           "(b-1) * l = 7*6 = 42",
                           "g-1 = 63", 
                           "1078",
                           "N-1 = 1187"))
knitr::kable(t_met, caption = "Sources of Variability for the Multi-environment trial")
```


#### Marginal means 

```{r}
emmeans(m_multienv, ~ gen)[1:10]
emmeans(m_multienv, ~ gen, contr = list(c(1, -1, rep(0, n_distinct(df_multienv$gen)-2))))$contr
```

## Subsampling   

**Review: split-plots**

- Split-plot designs are multi-level designed experiments. 
- Randomization happens in different levels. 
- Split-plot designs can happen in blocked designs or in CRDs.
- Importance of accounting for the subsampling (df and variance)

### Example  

A split-plot experiment in three blocks. Whole-plot is ’management’, sub-plot is ’time’ of application, with two subsamples. The data are the heights, measured on two single-hill sampling units in each plot.

```{r}
df_subsample <- gomez.splitplot.subsample
str(df_subsample)
m_subsample <- lmer(height ~ manage * time + (1|rep/manage/time), data = df_subsample)

VarCorr(m_subsample)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
t_subs <- data.frame(Source = c("Block", 
                                "Management",
                                "Error(whole plot) or Mgmt(Block)",
                                "Time",
                                "Mgmt x Time",
                                "Time(Mgmt x Block)",
                                "Subsample aka Error",
                                "Total"),
                    df = c("b - 1 = 2",
                           "m - 1 = 7",
                           "(m-1) * b - 7 = 7*3 -7 = 14",
                           "t - 1 = 3", 
                           "7*3 = 21",
                           "(t-1) * 8 * 3 - 24= 72 - 24 = 48",
                           "(s-1) * 4 * 8 * 3 = 96",
                           "N-1 = 191"))
knitr::kable(t_subs, caption = "Sources of Variability for the Multi-environment trial")
```

```{r}
emmeans(m_subsample, ~manage, contr = list(c(1, -1, 0, 0, 0, 0, 0, 0)))$contr
emmeans(m_subsample, ~time, contr = list(c(1, -1, 0, 0)))$contr
```

